---
title: 'STA 035C Homework 6: R Part'
author: "Yuan Xie"
date: "2025-05-29"
output: 
  pdf_document:
    keep_tex: no
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE)
```

## Import Environment and Dataset
```{r}
# import the tidyverse environment
library(tidyverse)
library(MASS)
```

---

## Problem 1
### (a)(iii)
```{r}
# input these data points
X <- matrix(c(
  2, 3,
  2, 5,
  3, 6,
  4, 5,
  5, 8,
  6, 10,
  6, 7,
  7, 11,
  7, 9,
  8, 12
), ncol = 2, byrow = TRUE)

# use prcomp to execute PCA
pca_result <- prcomp(X, center = TRUE, scale. = FALSE)

# output the first principal component
u1 <- pca_result$rotation[,1]
u1
```
**Interpretation:** The first principal component $\textbf{u}_1$ is (0.5856467, 0.8105665).

### (c)(i)
```{r}
# generate the data points
set.seed(111)
n <- 100
p <- 25
mean1 <- c(-2.5, -2.5, rep(0, p-2))
mean2 <- c(4, rep(0, p-1))
mean3 <- c(2, 3.5, -1, rep(0, p-3))
mean4 <- c(2, 3.3, 1, rep(0, p-3))
# X1 <- matrix(rnorm(n*p, mean=mean1, sd=1.0), n, p)
# X2 <- matrix(rnorm(n*p, mean=mean2, sd=1.0), n, p)
# X3 <- matrix(rnorm(n*p, mean=mean3, sd=1.0), n, p)
# X4 <- matrix(rnorm(n*p, mean=mean4, sd=1.0), n, p)
X1 <- MASS::mvrnorm(n, mean1, diag(1, nrow = p, ncol = p))
X2 <- MASS::mvrnorm(n, mean2, diag(1, nrow = p, ncol = p))
X3 <- MASS::mvrnorm(n, mean3, diag(1, nrow = p, ncol = p))
X4 <- MASS::mvrnorm(n, mean4, diag(1, nrow = p, ncol = p))
# Combine a "mixture" of 30 + 30 + 20 + 20 = 100 obs from 4 subgroups
X <- rbind(X1[1:30,], X2[1:30,], X3[1:20,], X4[1:20,])

# PCA using prcomp()
pca_result <- prcomp(X, center = TRUE, scale. = FALSE)

# First principal component direction
u1 <- pca_result$rotation[, 1]
u1

# Project data onto u1
proj_scores <- as.vector((scale(X, center = TRUE, scale = FALSE)) %*% u1)

# Compute variance along u1 direction
directional_variance <- mean(proj_scores^2)
directional_variance
```
**Interpretation:** The first principal component is shown above, which is very long. The directional variance along it is 23.23383.

### (c)(ii)
```{r}
sdev <- pca_result$sdev

# variance
var_explained <- sdev^2
var_explained_ratio <- var_explained / sum(var_explained)

# make a tiny data frame
scree_df <- tibble(
  PC = factor(1:length(var_explained_ratio)),
  VarianceExplained = var_explained_ratio
)

# draw Scree Plot
ggplot(scree_df, aes(x = PC, y = VarianceExplained, group = 1)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1, color = "red") +
  labs(title = "Scree Plot from PCA Using prcomp()",
       x = "Principal Component",
       y = "Proportion of Variance Explained")
```
**Interpretation:** The scree plot is shown above.

### (d)(i)
```{r}
# generate the data points
set.seed(111)
n <- 100
p <- 25
mean1 <- c(-2.5, -2.5, rep(0, p-2))
mean2 <- c(4, rep(0, p-1))
mean3 <- c(2, 3.5, -1, rep(0, p-3))
mean4 <- c(2, 3.3, 1, rep(0, p-3))
# X1 <- matrix(rnorm(n*p, mean=mean1, sd=1.0), n, p)
# X2 <- matrix(rnorm(n*p, mean=mean2, sd=1.0), n, p)
# X3 <- matrix(rnorm(n*p, mean=mean3, sd=1.0), n, p)
# X4 <- matrix(rnorm(n*p, mean=mean4, sd=1.0), n, p)
X1 <- MASS::mvrnorm(n, mean1, diag(1, nrow = p, ncol = p))
X2 <- MASS::mvrnorm(n, mean2, diag(1, nrow = p, ncol = p))
X3 <- MASS::mvrnorm(n, mean3, diag(1, nrow = p, ncol = p))
X4 <- MASS::mvrnorm(n, mean4, diag(1, nrow = p, ncol = p))
# Combine a "mixture" of 30 + 30 + 20 + 20 = 100 obs from 4 subgroups
X <- rbind(X1[1:30,], X2[1:30,], X3[1:20,], X4[1:20,])

# PCA using prcomp()
pca_result <- prcomp(X, center = TRUE, scale. = FALSE)
PCs <- as_tibble(pca_result$x[, 1:2])

set.seed(123)  # set reproducibility
km_orig <- kmeans(X, centers = 4, nstart = 25)

PCs_with_orig_cluster <- PCs %>%
  mutate(cluster_orig25D = factor(km_orig$cluster))

plot1 <- ggplot(PCs_with_orig_cluster, aes(x = PC1, y = PC2, color = cluster_orig25D)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "K-Means (k = 4) on 25D Data (Visualized in PC1-PC2 Plane)",
       x = "PC1", y = "PC2", color = "Cluster")

plot1
```
**Interpretation:** The results when k = 4 on PC1-PC2 plane are shown above.

### (d)(ii)
```{r}
set.seed(123)
km_pca <- kmeans(PCs, centers = 4, nstart = 25)

PCs_with_pca_cluster <- PCs %>%
  mutate(cluster_PC = factor(km_pca$cluster))

# plot a graph
plot2 <- ggplot(PCs_with_pca_cluster, aes(x = PC1, y = PC2, color = cluster_PC)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "K-Means (k = 4) on PC1-PC2 Projected Data",
       x = "PC1", y = "PC2", color = "Cluster")
plot2
```
**Interpretation:** The data has been projected onto the first two principal components. I have run k-means clustering with k = 4 on these 2D projected points. The separation between points becomes more apparent.

### (d)(iii)
```{r}
set.seed(123)
km_k3 <- kmeans(PCs, centers = 3, nstart = 25)

PCs_k3 <- PCs %>%
  mutate(cluster_k3 = factor(km_k3$cluster))

# plot the graph
ggplot(PCs_k3, aes(x = PC1, y = PC2, color = cluster_k3)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "K-Means Clustering with k = 3",
    subtitle = "Performed on PC1-PC2 Projected Data",
    x = "PC1", y = "PC2", color = "Cluster"
  ) 

set.seed(123)
km_k5 <- kmeans(PCs, centers = 5, nstart = 25)

PCs_k5 <- PCs %>%
  mutate(cluster_k5 = factor(km_k5$cluster))

# plot the graph
ggplot(PCs_k5, aes(x = PC1, y = PC2, color = cluster_k5)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "K-Means Clustering with k = 5",
    subtitle = "Performed on PC1-PC2 Projected Data",
    x = "PC1", y = "PC2", color = "Cluster"
  )
```
**Interpretation:** We did k-means clustering on the PC1-PC2 projected data when k = 3, k = 4, and k = 5. When the value of k increases, the number of clusters will increase. Some original entire clusters will be divided into two or more clusters.

---

## Problem 2
### (c)(ii)
```{r}
points <- matrix(c(
  2, 4,   # z1
  3, 3,   # z2
  3, 6,   # z3
  5, 15,  # z4
  6, 12,  # z5
  4, 2,   # z6
  10, 22, # z7
  11, 20, # z8
  12, 25, # z9
  8, 16   # z10
), ncol = 2, byrow = TRUE)

colnames(points) <- c("x", "y")
rownames(points) <- paste0("z", 1:10)

# C1: z1, z6
# C2: z2, z3, z5
# C3: z4, z7, z8, z9, z10
initial_clusters <- list(
  C1 = c(1, 6),
  C2 = c(2, 3, 5),
  C3 = c(4, 7, 8, 9, 10)
)

compute_centroid <- function(indices) {
  colMeans(points[indices, , drop = FALSE])
}

centroids <- t(sapply(initial_clusters, compute_centroid))
centroids

assign_to_nearest <- function(pt, centroids) {
  dists <- apply(centroids, 1, function(center) sum((pt - center)^2))
  which.min(dists)
}

new_cluster_ids <- apply(points, 1, assign_to_nearest, centroids = centroids)
cluster_names <- names(initial_clusters)[new_cluster_ids]

data.frame(Point = rownames(points), AssignedTo = cluster_names)
```
**Interpretation:** Each $z_i$ is reassigned to the closest centroid.

### (c)(iii)
```{r}
# the first update
updated_clusters <- list(
  C1 = which(cluster_names == "C1"),
  C2 = which(cluster_names == "C2"),
  C3 = which(cluster_names == "C3")
)

centroids_2 <- t(sapply(updated_clusters, function(indices) {
  colMeans(points[indices, , drop = FALSE])
}))
centroids_2

new_cluster_ids_2 <- apply(points, 1, assign_to_nearest, centroids = centroids_2)
cluster_names_2 <- names(updated_clusters)[new_cluster_ids_2]

data.frame(Point = rownames(points),
           ClusterAfterUpdate1 = cluster_names,
           ClusterAfterUpdate2 = cluster_names_2)
```
**Interpretation:** The centroids are recomputed and reassigned. The data frame shows the comparison.

### (d)
```{r}
points <- matrix(c(
  2, 4,   # z1
  3, 3,   # z2
  3, 6,   # z3
  5, 15,  # z4
  6, 12,  # z5
  4, 2,   # z6
  10, 22, # z7
  11, 20, # z8
  12, 25, # z9
  8, 16   # z10
), ncol = 2, byrow = TRUE)

colnames(points) <- c("x", "y")
rownames(points) <- paste0("z", 1:10)

set.seed(123) # keep the reproducibility

results <- list()

for (i in 1:5) {
  km <- kmeans(points, centers = 3, nstart = 1)
  results[[i]] <- list(
    clusters = km$cluster,
    tot_withinss = km$tot.withinss
  )
}

for (i in 1:5) {
  cat("Run", i, "\n")
  cat("Cluster assignment:", results[[i]]$clusters, "\n")
  cat("Total within-cluster sum of squares:", results[[i]]$tot_withinss, "\n\n")
}
```
**Interpretation:** The final solution looks different. The cluster assignment results are all different. However, the total within-cluster sum of squares keeps constant, which is always 38.75.

---

## Problem 3
### (c)
```{r}
points <- matrix(c(
  2, 4,   # z1
  3, 3,   # z2
  3, 6,   # z3
  5, 15,  # z4
  6, 12,  # z5
  4, 2,   # z6
  10, 22, # z7
  11, 20, # z8
  12, 25, # z9
  8, 16   # z10
), ncol = 2, byrow = TRUE)

rownames(points) <- paste0("z", 1:10)
colnames(points) <- c("x", "y")

dists <- dist(points)

hc <- hclust(dists, method = "complete")

plot(hc, labels = rownames(points), main = "Dendrogram (Complete Linkage)",
     xlab = "Points", ylab = "Height")
abline(h = 12, col = "red", lty = 2)

cut3 <- cutree(hc, k = 3)
cut3

merge_info <- as.data.frame(hc$merge)
merge_info$height <- hc$height
head(merge_info)

set.seed(123)
kmeans_result <- kmeans(points, centers = 3, nstart = 10)
kmeans_result$cluster

data.frame(Point = rownames(points),
           Hierarchical = cut3,
           KMeans = kmeans_result$cluster)

cutree(hc, k = 3)
split(rownames(points), cutree(hc, k = 3))
```
**Interpretation:** Points $z_1$ and $z_2$ merge first because their merging height is the lowest in the dendrogram. Before a “major” grouping forms, about 7 merges have occurred. The dendrogram is shown above. When k = 3, the clusters in the dendrograms are different from the result with k-means clustering. $z_{10}$ is assigned to the $C_2$ cluster with hierarchical clustering, but it is assigned to $C_3$ cluster with k-means clustering.








