---
title: 'STA 035C Homework 2: R Part'
author: "Yuan Xie"
date: "2025-04-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE)
```

## Import Environment and Dataset

```{r}
# import the tidyverse environment
library(tidyverse)
```

---

## Problem 3
### (c)
#### (i)

```{r}
# ensure the reproducibility
set.seed(123)
# dataset of size n = 50
n <- 50
# in a uniform distribution
X <- runif(n, min = 0, max = 30)
# the function of p
p <- 1 / (1 + exp(-(-5 + 0.25 * X)))
# generate Y
Y <- rbinom(n, size = 1, prob = p)
Y
```

#### (ii)

```{r}
# construct the logistic model
model <- glm(Y ~ X, family = binomial)
model
```

#### (iii)

```{r}
# summary() the model and pull the coefficients out
summary(model)$coefficients
```
**Answer:** The estimated intercept -9.40 and estimated slope 0.46 are different from the true parameters -5 and 0.25. They are not very close. However, it is normal to have some random variation in a small sample size.

#### (iv)

```{r}
# make a tiny data frame of Y against X
df <- data.frame(X = X, Y = Y)

# put X and Y into the tiny data frame
x_vals <- seq(0, 30, length.out = 300)
predicted_probs <- predict(model, newdata = data.frame(X = x_vals), type = "response")
fitted_df <- data.frame(X = x_vals, Prob = predicted_probs)

# plot the fitted logistic curve and the data points
df %>% ggplot(aes(x = X, y = Y)) +
  geom_point(color = "black", size = 2, alpha = 0.7) +  # actual data points
  geom_line(data = fitted_df, aes(x = X, y = Prob), color = "red", linewidth = 1.2) +  # fitted curve
  labs(title = "Logistic Regression Fit",
       x = "X",
       y = "Probability / Outcome")
```

#### (v)

```{r}
# predict probabilities for all X values
predicted_probs <- predict(model, type = "response")

# threshold 0.4
pred_class_0.4 <- ifelse(predicted_probs >= 0.4, 1, 0)
conf_matrix_0.4 <- table(Predicted = pred_class_0.4, Actual = Y)

# threshold 0.5
pred_class_0.5 <- ifelse(predicted_probs >= 0.5, 1, 0)
conf_matrix_0.5 <- table(Predicted = pred_class_0.5, Actual = Y)

# show the two confusion matrices
conf_matrix_0.4
conf_matrix_0.5
```
**Answer:** When using 0.4, the model is more lenient. There are fewer false negatives, but more false positives. When using 0.5, the model is stricter. There are fewer false positives, but more false negatives.

#### (vi)

**Answer:** With a greater sample size, the estimates of the intercept and slope will be closer to the true values. The logistic curve will fit the data better.

---

## Problem 4
### (a)
#### (i)

```{r}
# input the data into a tiny data frame
df <- data.frame(
  X1 = c(1.5, 2.0, 2.3, 3.0, 3.5, 4.1, 4.4, 5.2, 5.4),
  X2 = c(25, 50, 55, 60, 80, 75, 120, 100, 110),
  C = factor(c(1,1,1, 2,2,2, 3,3,3))
)

# calculate each class mean vector
mu1 <- colMeans(df[df$C == 1, 1:2])
mu2 <- colMeans(df[df$C == 2, 1:2])
mu3 <- colMeans(df[df$C == 3, 1:2])
# show the mean vectors
mu1
mu2
mu3

# calculate the common covariance matrix with 2 degrees of freedom
matrix_cov <- (
  cov(df[df$C == 1, 1:2]) * 2 +
  cov(df[df$C == 2, 1:2]) * 2 +
  cov(df[df$C == 3, 1:2]) * 2
) / (9 - 3)

# show the common covariance matrix
matrix_cov
```

#### (ii)

```{r}
# use solve() to get the inverse of the matrix
Sigma_inv <- solve(matrix_cov)
# show the inverse matrix
Sigma_inv

# make the linear discriminant functions for the three cases
delta1 <- function(x1, x2) {
  x <- c(x1, x2)
  as.numeric(t(x) %*% Sigma_inv %*% mu1 - 0.5 * t(mu1) %*% Sigma_inv %*% mu1)
}

delta2 <- function(x1, x2) {
  x <- c(x1, x2)
  as.numeric(t(x) %*% Sigma_inv %*% mu2 - 0.5 * t(mu2) %*% Sigma_inv %*% mu2)
}

delta3 <- function(x1, x2) {
  x <- c(x1, x2)
  as.numeric(t(x) %*% Sigma_inv %*% mu3 - 0.5 * t(mu3) %*% Sigma_inv %*% mu3)
}

a1 <- Sigma_inv %*% mu1
a2 <- Sigma_inv %*% mu2
a3 <- Sigma_inv %*% mu3

c1 <- -0.5 * t(mu1) %*% Sigma_inv %*% mu1
c2 <- -0.5 * t(mu2) %*% Sigma_inv %*% mu2
c3 <- -0.5 * t(mu3) %*% Sigma_inv %*% mu3

# show the 3 discriminant functions
cat("delta_1(x) = ", a1[1], "* x1 +", a1[2], "* x2 +", c1, "\n")
cat("delta_2(x) = ", a2[1], "* x1 +", a2[2], "* x2 +", c2, "\n")
cat("delta_3(x) = ", a3[1], "* x1 +", a3[2], "* x2 +", c3, "\n")
```

#### (iii)

```{r}
# define the range of x and y axes here
x1_seq <- seq(1, 6, length.out = 200)
x2_seq <- seq(20, 130, length.out = 200)
# make a grid to store the data
grid <- expand.grid(x1 = x1_seq, x2 = x2_seq)

# predict the data points in the grid
grid <- grid %>%
  # process row by row
  rowwise() %>%
  mutate(
    d1 = delta1(x1, x2),
    d2 = delta2(x1, x2),
    d3 = delta3(x1, x2),
    pred = which.max(c(d1, d2, d3))
  ) %>%
  # delete the rowwise structure
  ungroup() %>%
  # add the class title
  mutate(pred = factor(pred, labels = c("Light", "Moderate", "Heavy")))

# change it to factors
df$C <- factor(df$C, labels = c("Light", "Moderate", "Heavy"))

# plot the graph
ggplot() +
  geom_tile(data = grid, aes(x = x1, y = x2, fill = pred), alpha = 0.5) +
  geom_point(data = df, aes(x = X1, y = X2, color = C), size = 3) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "pink"),
                    name = "Predicted Class") +
  scale_color_manual(values = c("blue", "green", "red"),
                     name = "True Class") +
  labs(x = "Screen Time (hrs/day)",
       y = "Call Time (mins/week)",
       title = "LDA Decision Boundaries")
```
**Answer:** The decision boundaries are determined when the discriminant functions are equal. The boundaries will be linear.

### (b)
#### (i)

```{r}
# the same as the last question
df$C <- factor(df$C, labels = c("Light", "Moderate", "Heavy"))

# use nnet::multinom to fit the data
library(nnet)
model_mlr <- multinom(C ~ X1 + X2, data = df)

# the same as the last question
x1_seq <- seq(1, 6, length.out = 200)
x2_seq <- seq(20, 130, length.out = 200)
grid <- expand.grid(X1 = x1_seq, X2 = x2_seq)

grid$pred <- predict(model_mlr, newdata = grid)

# the same as the last question
ggplot() +
  geom_tile(data = grid, aes(x = X1, y = X2, fill = pred), alpha = 0.3) +
  geom_point(data = df, aes(x = X1, y = X2, color = C), size = 3) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "pink"),
                    name = "Predicted Class") +
  scale_color_manual(values = c("blue", "green", "red"),
                     name = "True Class") +
  labs(x = "Screen Time (hrs/day)",
       y = "Call Time (mins/week)",
       title = "Multinomial Logistic Regression Decision Boundaries")
```
**Answer:** The boundary of the multiple logistic regression is not linear compared to LDA.

#### (ii)

```{r}
lda_pred_class <- {
  d1 <- delta1(3.1, 90)
  d2 <- delta2(3.1, 90)
  d3 <- delta3(3.1, 90)
  c("Light", "Moderate", "Heavy")[which.max(c(d1, d2, d3))]
}

new_user <- data.frame(X1 = 3.1, X2 = 90)
mlr_pred_class <- predict(model_mlr, newdata = new_user)

cat("LDA prediction: ", lda_pred_class, "\n")
cat("Multinomial logistic prediction: ", mlr_pred_class, "\n")
```
**Answer:** LDA assigns to moderate. Mutinomial logistic prediction assigns to light. The predictions are different.

### (c)

I have no idea...









