---
title: 'STA 035C Homework 4: R Part'
author: "Yuan Xie"
date: "2025-05-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE)
```

## Import Environment and Dataset
```{r}
# import the tidyverse environment
library(tidyverse)
```

---

## Problem 1
### (a)
```{r}
# generate the data
set.seed(123)
n <- 100
X <- rnorm(n)
log_odds <- -2 - 3*X
probs <- 1 / (1 + exp(-log_odds))
Y <- rbinom(n, 1, probs)
df <- data.frame(X, Y=as.factor(Y))

# draw the scatterplot
df %>% ggplot(aes(x = X, y = Y, color = Y)) +
  geom_jitter(height = 0.05, width = 0, size = 2) +
  labs(title = "Scatterplot of X vs. Y (0/1 Classes)",
       x = "X", 
       y = "Y (Class 0 or 1)")
```

### (b)
```{r}
set.seed(1)
n_sim <- 100
errors <- numeric(n_sim)
for (i in 1:n_sim) {
  # randomly divide the data
  idx <- sample(1:nrow(df), size = nrow(df) / 2)
  train <- df[idx, ]
  test <- df[-idx, ]
  
  # fit the logistic regression model
  model <- glm(Y ~ X, data = train, family = binomial)
  
  # predict the probability
  probs <- predict(model, newdata = test, type = "response")
  preds <- ifelse(probs > 0.5, "1", "0")
  
  # calculate the error rate
  error <- mean(preds != test$Y)
  errors[i] <- error
}

error_df <- data.frame(ErrorRate = errors)
error_df %>% ggplot(aes(x = ErrorRate)) +
  geom_histogram(binwidth = 0.02, fill = "skyblue", color = "black") +
  labs(title = "Classification Error (100 Splits)",
       x = "Classification Error Rate",
       y = "Count")

# calculate the mean, min and max
mean_err <- mean(errors)
min_err <- min(errors)
max_err <- max(errors)
cat("Mean error:", mean_err, "\n")
cat("Min error:", min_err, "\n")
cat("Max error:", max_err, "\n")
```
**Interpretation:** Mean error: 0.1506; Min error: 0.08; Max error: 0.22 

### (c)
```{r}
n <- nrow(df)
loocv_errors <- numeric(n)

for (i in 1:n) {
  train <- df[-i, ]
  test <- df[i, , drop = FALSE]
  
  # fit the logistic regression model
  model <- glm(Y ~ X, data = train, family = binomial)
  prob <- predict(model, newdata = test, type = "response")
  pred <- ifelse(prob > 0.5, "1", "0")
  
  loocv_errors[i] <- pred != test$Y
}

# LOOCV error rate
loocv_error_rate <- mean(loocv_errors)

# compare the output
cat("LOOCV Error Rate:", loocv_error_rate, "\n")
cat("Mean Error from (b):", mean(errors), "\n")
cat("Variance of Error from (b):", var(errors), "\n")
```
**Interpretation:** Compared to the error rate of random 50/50 splits, the LOOCV method provided a slightly higher error but with greater stability.

### (d)
```{r}
set.seed(1)

# set the values of k
k_values <- c(2, 3, 4, 5, 10, 20)
cv_errors <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  n <- nrow(df)
  folds <- sample(rep(1:k, length.out = n))
  fold_errors <- numeric(k)
  
  for (j in 1:k) {
    test_idx <- which(folds == j)
    train_data <- df[-test_idx, ]
    test_data <- df[test_idx, ]
    
    # set the logistic regression model
    model <- glm(Y ~ X, data = train_data, family = binomial)
    probs <- predict(model, newdata = test_data, type = "response")
    preds <- ifelse(probs > 0.5, "1", "0")
    
    fold_errors[j] <- mean(preds != test_data$Y)
  }
  
  cv_errors[i] <- mean(fold_errors)
}

cv_df <- data.frame(k = k_values, ErrorRate = cv_errors)
print(cv_df)

# plot the point and line graph
cv_df %>% ggplot(aes(x = k, y = ErrorRate)) +
  geom_line() +
  geom_point(size = 3) +
  labs(title = "k-Fold CV Classification Error",
       x = "Number of Folds (k)",
       y = "Classification Error Rate")
```
**Interpretation:**  The resulting classification error rates are shown in the tibble above. The highest classification error occurred at k = 3, with an error rate above 18%. The lowest error was observed at both k = 2 and k = 4, with error rates close to 15%. After k becomes equal or larger than 5, the error stabilizes around 16%.

### (e)
```{r}
######### code in this part is generally the same except adding the quadratic term

# generate data with quadratic term X^2
set.seed(123)
n <- 100
X <- rnorm(n)
X2 <- X^2
log_odds <- -2 - 3 * X + X^2
probs <- 1 / (1 + exp(-log_odds))
Y <- rbinom(n, 1, probs)
df2 <- data.frame(X = X, X2 = X2, Y = as.factor(Y))


### repeat (b)
set.seed(1)
n_sim <- 100
errors2 <- numeric(n_sim)

for (i in 1:n_sim) {
  idx <- sample(1:nrow(df2), size = nrow(df2) / 2)
  train <- df2[idx, ]
  test <- df2[-idx, ]
  
  model <- suppressWarnings(glm(Y ~ X + X2, data = train, family = binomial))
  probs <- predict(model, newdata = test, type = "response")
  preds <- ifelse(probs > 0.5, "1", "0")
  
  error <- mean(preds != test$Y)
  errors2[i] <- error
}

# draw the histogram
data.frame(ErrorRate = errors2) %>%
  ggplot(aes(x = ErrorRate)) +
  geom_histogram(binwidth = 0.02, fill = "skyblue", color = "black") +
  labs(title = "Classification Error with Quadratic Term (100 Splits)",
       x = "Classification Error Rate",
       y = "Count")

# output summary
cat("Mean error (quad):", mean(errors2), "\n")
cat("Min error (quad):", min(errors2), "\n")
cat("Max error (quad):", max(errors2), "\n")




### repeat (c)
n <- nrow(df2)
loocv_errors2 <- numeric(n)

for (i in 1:n) {
  train <- df2[-i, ]
  test <- df2[i, , drop = FALSE]
  
  model <- suppressWarnings(glm(Y ~ X + X2, data = train, family = binomial))
  prob <- predict(model, newdata = test, type = "response")
  pred <- ifelse(prob > 0.5, "1", "0")
  
  loocv_errors2[i] <- pred != test$Y
}

# output summary
cat("LOOCV Error Rate (quad):", mean(loocv_errors2), "\n")
cat("Mean Error from (b):", mean(errors2), "\n")
cat("Variance of Error from (b):", var(errors2), "\n")




### repeat (d)
set.seed(1)
k_values <- c(2, 3, 4, 5, 10, 20)
cv_errors2 <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  k <- k_values[i]
  folds <- sample(rep(1:k, length.out = n))
  fold_errors <- numeric(k)
  
  for (j in 1:k) {
    test_idx <- which(folds == j)
    train_data <- df2[-test_idx, ]
    test_data <- df2[test_idx, ]
    
    model <- suppressWarnings(glm(Y ~ X + X2, data = train_data, family = binomial))
    probs <- predict(model, newdata = test_data, type = "response")
    preds <- ifelse(probs > 0.5, "1", "0")
    
    fold_errors[j] <- mean(preds != test_data$Y)
  }
  
  cv_errors2[i] <- mean(fold_errors)
}

cv_df2 <- data.frame(k = k_values, ErrorRate = cv_errors2)
print(cv_df2)

# plot the point and line graph
cv_df2 %>% ggplot(aes(x = k, y = ErrorRate)) +
  geom_line() +
  geom_point(size = 3) +
  labs(title = "k-Fold CV with Quadratic Term",
       x = "Number of Folds (k)",
       y = "Classification Error Rate")
```
**Interpretation:** Repeat (b): Mean error: 0.151; Min error: 0.08; Max error: 0.24. Repeat (c): Compared to the error rate of random 50/50 splits, the LOOCV method provided a slightly lower error but with greater stability. Repeat (c): The resulting classification error rates are shown in the tibble above. The highest classification error occurred at k = 4, with an error rate 16%. The lowest error was observed at both k = 3, with an error rate about 13.1%. When k is between 5 and 10, the error stabilizes around 15%. When k is between 10 and 20, the error rate gradually decreases from 15% to 14%.

---

## Problem 2
### (c)
```{r}
coverage <- numeric(1000)

# set a for loop for the seeds
for (k in 1:1000) {
  set.seed(k)
  x <- rnorm(100)
  y <- -1 + 2 * x + rnorm(100)
  
  model <- lm(y ~ x)
  conf_int <- confint(model, level = 0.95)
  lower <- conf_int["x", 1]
  upper <- conf_int["x", 2]
  
  # whether it includes true beta = 2
  coverage[k] <- (lower <= 2) & (2 <= upper)
}

# calculate the coverage rate
coverage_rate <- mean(coverage)
cat("95% CI coverage rate for beta = 2:", coverage_rate, "\n")
```
**Interpretation:** The fraction of these covering the true slope is 0.951 and it is quite close to 0.95.

### (d)
```{r}
set.seed(1)

n_datasets <- 1000
n_boot <- 1000
n <- 100
true_beta1 <- 2
bootstrap_coverage <- numeric(n_datasets)

for (k in 1:n_datasets) {
  set.seed(k)
  x <- rnorm(n)
  y <- -1 + 2 * x + rnorm(n)
  df <- data.frame(x = x, y = y)

  # make the model
  model <- lm(y ~ x, data = df)
  beta1_hat <- coef(model)[["x"]]
  
  # bootstrap and repeat 1000 times
  beta1_star <- numeric(n_boot)
  for (b in 1:n_boot) {
    boot_idx <- sample(1:n, size = n, replace = TRUE)
    boot_df <- df[boot_idx, ]
    boot_model <- lm(y ~ x, data = boot_df)
    beta1_star[b] <- coef(boot_model)[["x"]]
  }
  
  # bootstrap standard deviation and make normal-based CI
  boot_se <- sd(beta1_star)
  
  # show the first 10 standard deviations
  if (k <= 10) {
    cat("Simulation", k, "- Bootstrap SD:", boot_se, "\n")
  }
  
  ci_lower <- beta1_hat - qnorm(0.975) * boot_se
  ci_upper <- beta1_hat + qnorm(0.975) * boot_se
  
  # whether it contains true beta = 2
  bootstrap_coverage[k] <- (ci_lower <= true_beta1) & (true_beta1 <= ci_upper)
}

# output coverage rate
boot_coverage_rate <- mean(bootstrap_coverage)
cat("Bootstrap 95% CI coverage rate for beta = 2:", boot_coverage_rate, "\n")
```
**Interpretation:** Now, the fraction of these covering the true slope is 0.932, which is slightly smaller than 95%. However, the value in (b) is 0.951, which is slightly bigger than 95%. The bootstrap confidence intervals may not be normally distributed.

### (d)
```{r}
##### the code of this part is similar to (c) and (d)

### (e/c) Standard 95% CI under nonlinear model
coverage <- numeric(1000)

for (k in 1:1000) {
  set.seed(k)
  x <- rnorm(100)
  y <- -1 + 2 * x - x^2 + rnorm(100)  # nonlinear data
  
  model <- lm(y ~ x)
  conf_int <- confint(model, level = 0.95)
  lower <- conf_int["x", 1]
  upper <- conf_int["x", 2]
  
  # whether it contains true beta = 2
  coverage[k] <- (lower <= 2) & (2 <= upper)
}

# final coverage rate
coverage_rate <- mean(coverage)
cat("Standard 95% CI coverage rate for beta = 2 (nonlinear):", coverage_rate, "\n")


### (e/d) Bootstrap normal CI under nonlinear model
set.seed(1)

n_datasets <- 1000
n_boot <- 1000
n <- 100
true_beta1 <- 2
bootstrap_coverage <- numeric(n_datasets)

for (k in 1:n_datasets) {
  set.seed(k)
  x <- rnorm(n)
  y <- -1 + 2 * x - x^2 + rnorm(n)  # nonlinear data
  df <- data.frame(x = x, y = y)

  model <- lm(y ~ x, data = df)
  beta1_hat <- coef(model)[["x"]]
  
  beta1_star <- numeric(n_boot)
  for (b in 1:n_boot) {
    boot_idx <- sample(1:n, size = n, replace = TRUE)
    boot_df <- df[boot_idx, ]
    boot_model <- lm(y ~ x, data = boot_df)
    beta1_star[b] <- coef(boot_model)[["x"]]
  }
  
  # standard deviation and CI
  boot_se <- sd(beta1_star)
  if (k <= 10) {
    cat("Simulation", k, "- Bootstrap SD:", boot_se, "\n")
  }
  
  ci_lower <- beta1_hat - qnorm(0.975) * boot_se
  ci_upper <- beta1_hat + qnorm(0.975) * boot_se
  
  bootstrap_coverage[k] <- (ci_lower <= true_beta1) & (true_beta1 <= ci_upper)
}

# final bootstrap coverage
boot_coverage_rate <- mean(bootstrap_coverage)
cat("Bootstrap 95% CI coverage rate for beta = 2 (nonlinear):", boot_coverage_rate, "\n")
```
**Interpretation:** The repeated (c) and (d) values become obviously smaller (0.683 and 0.894). This is because the real model is actually linear and the model we chose is wrong. Bootstrap performs better in this case because it resamples based on the data generated. It partly relieves the problem of wrong model. However, both the fractions here are smaller than 95%.

---

## Problem 3
### (b)(i)
```{r}
# input the dataset and relative packages
credit <- read_csv("Credit.csv")
head(credit)

# for combinations()
library(gtools)  

# set the response variable
response_var <- "Balance"
predictors <- setdiff(names(credit), response_var)

# store all the RSS and k values
all_models <- list()

for (k in 1:length(predictors)) {
  combos <- combinations(n = length(predictors), r = k, v = predictors)
  
  for (i in 1:nrow(combos)) {
    vars <- combos[i, ]
    formula <- as.formula(paste(response_var, "~", paste(vars, collapse = "+")))
    fit <- lm(formula, data = credit)
    rss <- sum(residuals(fit)^2)
    
    all_models[[length(all_models) + 1]] <- data.frame(
      k = k,
      RSS = rss,
      Variables = paste(vars, collapse = ", ")
    )
  }
}

# store these data into one data frame
all_models_df <- bind_rows(all_models)

# plot the graph
ggplot(all_models_df, aes(x = k, y = RSS)) +
  geom_jitter(width = 0.2, height = 0, color = "blue", alpha = 0.6) +
  labs(
    title = "RSS for All Subsets vs. Subset Size k",
    x = "Subset Size k",
    y = "Residual Sum of Squares (RSS)"
  )
```
**Interpretation:** This is the scatter plot of RSS versus subset size k for all subsets.

### (b)(ii)
```{r}
library(leaps)

response_var <- "Balance"
predictors <- setdiff(names(credit), response_var)

# make the formula
formula <- as.formula(paste(response_var, "~", paste(predictors, collapse = "+")))

# do the best subset selection
regfit.full <- regsubsets(formula, data = credit, nvmax = length(predictors))

# get the RSS data
reg_summary <- summary(regfit.full)

# make a data frame to plot the graph
rss_df <- data.frame(
  k = 1:length(reg_summary$rss),
  RSS = reg_summary$rss
)

which_mat <- summary(regfit.full)$which[, -1]

# get the variable names 
var_names <- apply(which_mat, 1, function(row) {
  paste(names(row)[row], collapse = ", ")
})

rss_df_full <- rss_df %>%
  mutate(Variables = var_names)

# get the smallest RSS
rss_df_full %>%
  group_by(k) %>%
  arrange(k)
```
**Interpretation:** For each k, the subsets with the lowest RSS are shown in the data frame above.

### (b)(iii)
```{r}
best_index <- which.max(reg_summary$adjr2)

# choose the row
which_row <- reg_summary$which[best_index, ]

# choose the names of the predictors
best_variables <- names(which_row)[which_row]
best_variables <- setdiff(best_variables, "(Intercept)")

cat("Adjusted R^2 is maximized at subset size k =", best_index, "\n")
cat("Best subset variables:", paste(best_variables, collapse = ", "))
```
**Interpretation:** The adjusted R-squared is maximized when k = 7. The best subset variables are: Income, Limit, Rating, Cards, Age, OwnYes, StudentYes.

### (b)(iv)
```{r}
set.seed(1) # reproducibility

predict_regsubsets <- function(object, newdata, id) {
  mat <- model.matrix(formula, newdata)
  coefi <- coef(object, id = id)
  vars <- names(coefi)
  mat[, vars] %*% coefi
}

# 5-fold CV set
k <- 5
folds <- sample(1:k, nrow(credit), replace = TRUE)
cv_errors <- matrix(NA, k, length(predictors))

# train and test for every fold
for (j in 1:k) {
  train <- credit[folds != j, ]
  test  <- credit[folds == j, ]
  
  regfit.k <- regsubsets(formula, data = train, nvmax = length(predictors))
  
  for (i in 1:length(predictors)) {
    pred <- predict_regsubsets(regfit.k, newdata = test, id = i)
    true <- test[[response_var]]
    cv_errors[j, i] <- mean((true - pred)^2)
  }
}

# mean MSE for each k
mean_cv_errors <- colMeans(cv_errors)
best_k_cv <- which.min(mean_cv_errors)

# best cross validation variables
coef_vec <- coef(regfit.full, id = best_k_cv)
best_cv_vars <- names(coef_vec)[-1]  # no intercept

cat("Best subset size by 5-fold CV is:", best_k_cv, "\n")
cat("Best subset variables by CV:", paste(best_cv_vars, collapse = ", "))
```
**Interpretation:** The answer here is different from the previous question. The best model by 5-fold cross-validation has 6 predicting variables. The best subset variables are: Income, Limit, Rating, Cards, Age, StudentYes. 

### (c)
```{r}
# Forward Stepwise Selection
regfit.fwd <- regsubsets(formula, data = credit, nvmax = length(predictors), method = "forward")
fwd_summary <- summary(regfit.fwd)

fwd_which <- fwd_summary$which[, -1]  # remove intercept
fwd_vars <- apply(fwd_which, 1, function(row) {
  paste(names(row)[row], collapse = ", ")
})

# store in a data frame
fwd_df <- data.frame(
  k = 1:length(fwd_vars),
  Variables = fwd_vars
)

# show the output
fwd_df
```
**Interpretation:** They are roughly the same, but when k = 4, "Rating" is included here instead of "Cards". Some small difference still exists.

---

## Problem 4
### (b)(a)
```{r}
# input the dataset and relative packages
college <- read_csv("College.csv") %>% select(-1)
head(college)

set.seed(1) # reproducibility
train_index <- sample(1:nrow(college), 0.7 * nrow(college))

# divide into two groups
train_data <- college[train_index, ]
test_data <- college[-train_index, ]
```
**Interpretation:** The dataset has been divided into a training set and a test set here.

### (b)(b)
```{r}
# fit the two linear models
lm_fit <- lm(Apps ~ ., data = train_data)
lm_preds <- predict(lm_fit, newdata = test_data)

# calculate the test MSE
test_mse <- mean((test_data$Apps - lm_preds)^2)
# output the test MSE
test_mse
```
**Interpretation:** The test MSE obtained is 1261630.

### (b)(c)
```{r}
suppressPackageStartupMessages(library(glmnet))

X_train <- model.matrix(Apps ~ ., data = train_data)[, -1]  # remove intercept
y_train <- train_data$Apps

X_test <- model.matrix(Apps ~ ., data = test_data)[, -1]
y_test <- test_data$Apps

# use cross-validation
set.seed(1)
cv_ridge <- cv.glmnet(
  X_train, y_train, 
  alpha = 0,                 # ridge regression
  standardize = TRUE        # standardization
)

# best fit lambda
best_lambda <- cv_ridge$lambda.min
ridge_fit <- glmnet(X_train, y_train, alpha = 0, lambda = best_lambda)

# predict based on the dataset
ridge_preds <- predict(ridge_fit, s = best_lambda, newx = X_test)

# calculate the ridge test MSE
ridge_test_mse <- mean((y_test - ridge_preds)^2)
ridge_test_mse
```
**Interpretation:** The ridge regression test MSE is 1121808.

### (b)(d)
```{r}
# Prepare x, y for glmnet (no intercept => '-1')
X_train <- model.matrix(Apps ~ . - 1, data = train_data)
y_train <- train_data$Apps

X_test <- model.matrix(Apps ~ . - 1, data = test_data)
y_test <- test_data$Apps

# Cross-validation to find optimal lambda
set.seed(1)
cv_lasso <- cv.glmnet(
  X_train, y_train,
  alpha = 1,                  # Lasso
  intercept = FALSE,
  standardize = TRUE
)

# Best lambda
best_lambda_lasso <- cv_lasso$lambda.min

# Fit Lasso with best lambda
lasso_fit <- glmnet(
  X_train, y_train,
  alpha = 1,
  lambda = best_lambda_lasso,
  intercept = FALSE,
  standardize = TRUE
)

# Print Lasso Coefficients
cat("Lasso Coeffs (lambda =", best_lambda_lasso, ")",
    as.matrix(coef(lasso_fit)), "\n")

# Predict and calculate test MSE
lasso_preds <- predict(lasso_fit, s = best_lambda_lasso, newx = X_test)
lasso_test_mse <- mean((y_test - lasso_preds)^2)
cat("Lasso Test MSE:", lasso_test_mse, "\n")

# Count non-zero coefficients (excluding intercept)
num_nonzero <- sum(as.matrix(coef(lasso_fit))[-1, ] != 0)
cat("Number of non-zero coefficients:", num_nonzero)
```
**Interpretation:** The Lasso regression test MSE is 1228976. The number of non-zero coefficient is 14.

### (b)(g)
**Answer:** The models are able to predict the number of applications. The three results are not far apart. The ordinary least squares regression has the highest test error (1261630). Ridge regression gets the lowest test MSE (1121808). Overall, the three models all perform well.














